Parity games in pushdown systems with parameters

### 2023-06-21 reachability/article.tex
up

% {\bf Context}

{\bf Games on pushdown systems} Pushdown systems can be used to model the behavior of recursive programs.
As a result, a variety of 
 program analysis questions can be reduced to decision problems for games on pushdown systems.
They have applications for instance in inter-procedural control-flow analysis of recursive programs \cite{esparza1999automata, reps2005weighted}.

@@

{\bf Parametric pushdown systems}
 Parametric pushdown systems provide a  formalism to
reason about 
recursive programs
making use of parametric constraints.
% Here, the stack can additionally be compared against parameters that can take unspecified word values. 
Recently, different variants of parametric pushdown systems have been introduced in the literature 
% [Hague’11]=[20], [Esparza, Ganty, Majumdar’13]=[15]
\cite{hague2011parameterised, esparza2016parameterized, fortin2017model} 
% In this work, following [21,20,15,27,12], we consider parametric pushdown systems
however as
parameterized asynchronous shared-memory systems.
These systems consist of a leader pushdown automaton
and arbitrarily many identical contributor pushdown systems, communicating via a shared memory
in the form of a register which can take finitely many values.
This variant have been shown to have applications
to the dataflow analysis of concurrent programs \cite{kahlon2008parameterization}. 
We consider here a different approach to extending pushdown systems with parameters, by allowing them to test equality of the stack content against parameters.
% Pushdown systems which can test the stack content against regular-expressions are equivalent to pushdown systems, so pushdown systems which can test the stack content against specific words are too.
Perhaps a similar model consists in 
pushdown systems 
with transitions that are conditioned by regular conditions on their stack content, which
can in particular
test the
stack content against
specific words. 
They can be used to ensure that some word over the stack alphabet appears in the configurations of a run, but only for a specific value (or, rather, specific regular languages).
They have \textcolor{red}{ phrases trop longues}
been used to establish that CTL$^*$ model checking remains decidable
when the formulas are allowed to include regular predicates on the stack content,
and to obtain 
model checking algorithms for LTL and CTL$^*$ model checking
for
pushdown systems \cite{finkel1997direct}.
Pushdown systems 
%with regularity condition on the stack
with transitions that are conditioned by regular conditions on their stack content
can be viewed as
a special case of stack automata as seen in \cite{hopcroft1969formal}.
Instead
of checking the stack content against regular conditions, or, in a more limited manner,
against specified word values,
we consider
checks against unspecified word values that can be instantiated using parameters.
This allows for example to ensure equality between two stack contents appearing in two distincts configurations in a run.

@@

% Hence we propose in this paper to extend pushdown systems with a set of  parameters.
%
% Compared to pebble word automata, instead of using pebbles
% as markings on their input, pebble pushdown systems have the ability to lift or drop pebbles
% on their universe of stack contents. Dropping a pebble records the current stack content with an available pebble until it is lifted. Lifting a pebble then makes it available again.

% A pebble automaton is a two-way finite state automaton that uses a fixed, finite number of
% pebbles that it can drop on, and lift from 
% words, using them as markers.
% Pebble automata recognize regular languages only, provided the life times of the pebbles, i.e. the times between dropping a pebble and lifting it again, are properly nested
% \cite{globerman1996complexity, engelfriet1999tree}.
% Automata with nested pebbles were also introduced for tree-walking automata. 
% It is known that tree-walking automata do not recognize all regular tree languages \cite{bojanczyk2008tree}, the main trouble being that the tree-walking automata get lost rather easily.
% Using pebbles is a remedy against getting lost along a tree, but  the unrestricted use of pebbles leads to a class of tree languages much larger than the regular tree languages, in fact to all tree languages in $\NSPACE(log ~ n)$.
% Thus, in both pebble word automata and pebble tree-walking automata, the placement of the pebble follows a strict stack discipline, and our model will follow this strict stack discipline too.
%
%
% One motivation to the study of pebble pushdown systems is their ability to compare a stack content against some later stack content.
% Similar type of storage for later comparison has already been used for instance in register pushdown systems. Introduced in \cite{murawski2017reachability}  as an extension of pushdown systems that can keep data values in a set of registers and a stack, register pushdown systems enable manipulation of data values from an infinite domain, and have applications for instance for malware detection and XML schema checking \cite{senda2021forward, senda2021ltl}.

Another closely related formalism is that of alternating tree automata with nested pebbles%, as we consider in more details in Section 4
. The approach of extending tree automata with pebbles was used in \cite{milo2000typechecking} to show that the XML typechecking problem is decidable, and in \cite{karhumaki2012jewels} to recognize first-order logic on trees. In both cases however, the input trees considered are finite, while 
simulation of plays in parametric pushdown graphs
requires the use of pebble alternating tree automata working on infinite trees
since the stack of a pushdown system is in general unbounded.

@@

We provide formal definitions of parametric pushdown systems, parametric pushdown reachability games
and parametric pushdown parity games in Section~\ref{section def ppda}. We give an overview of our contribution in Section~\ref{contribution ppda}. The contribution consists in
proving parametric pushdown reachability games
and parametric pushdown parity games belong to $(n+1)$-$\EXP$ in case the number of parameters
$n$ is fixed,
and providing a nonelementary lower bound for parametric pushdown reachability games in general.
We provide the proof of our result, which stretches along 
Section~\ref{lower bound PDA},
Section~\ref{pebbles},
% Section~\ref{ATWA and PATWA}, and
Section~\ref{section HPDA}.
In Section~\ref{discuss ppda} we close the chapter with a discussion about the methods used, with some directions for future work.
\fi

@@

By $\Z $ we denote the {\em integers} and by $\N=\{0,1,\ldots\}$ we denote the {\em naturals}.
% For every $a,b\in\Z$ with $a\leq b$ we define $[a,b]=\{k\in\Z\mid a\leq k\leq b\}$.
% For every $n\geq 1$ we define $n\Z=\{n\cdot z\mid z\in\Z\}$.
% For every number $n\in\N$ we define $\log(n)=\min\{i+1\mid i\in\N, n\leq 2^i\}$, which is the smallest number of bits necessary to write down $n$ in binary.
For every finite alphabet $A$, $A^*$ is the set of finite
words with letters in  $A$, $A^\omega$ is the set of infinite words with letters in $A$, and
%by
 $A^\infty = A^\omega \cup A^*$. We denote the empty word in $A^*$ by $\epsilon$.
%For all $a\in A$ and all $w\in A^*$ let $|w|_a$ denote the number of occurrences of the letter $a$ in $w$.
%
% For every finite set $M\subset\N\setminus\{0\}$ let $\LCM(M)=\min\{n\geq 1\mid \forall m\in M\setminus\{0\}: m|n\}$ denote the least common multiple of the elements in $M$. 
% For any $j \in \N$ let $\LCM(j)=\LCM([1,j])$ denote the least common multiple of the numbers $\{1,\ldots,j\}$.

 For any two sets $X$ and $S$, let $X^S$ denote the set of all functions from $S$ to $X$.
For any set $S$ let 
%$\mathscr{P}(S) = \{ X \mid X \subseteq S \}$
$\mathcal{P}(S) = \{ X \mid X \subseteq S \}$
denote the power set of $S$.
%
A {\em partial function} $f$ from a set $X$ to a set $Y$ is a function defined on a subset $C$ of $X$ (possibly $X$ itself) with output values in $Y$. We extend the domain of a partial function to the whole set
by considering it as returning the bottom element $\bot$ when it is undefined.

A {\em graph} $G=(V,E)$ is a finite set of graph vertices $V$ with a set of graph edges $E \subseteq V \times V$. 
% Two graphs $G = (V,E)$ and $G'=(V',E')$ with sets of graph vertices $V=V'=V_n=\{1,2,...,n\}$ are said to be {\em isomorphic} if there is a permutation $\rho$ of $V_n$ such that $(u,v)$ is in the set of graph edges $E$ iff $(\rho(u),\rho(v))$ is in the set of graph edges $E'$.
%%%
%
For any set $V$, and a given infinite sequence $\pi = v_0 v_1 \ldots \in V^\omega$ of elements of $V$, we define the set 
$\text{\textit{Inf}}(\pi)$ of elements occurring infinitely often in $\pi$ as
$\text{\textit{Inf}}(\pi) = \{ v \in V \mid \forall i, \exists j > i, ~ v_j = v \}$.

@@

%ù \subsection{Games}

@@

% \subsubsection{Arenas}

\begin{samepage}
An {\em arena} is a tuple $A=(S_0, S_1,{\rightarrow})$  which is composed of
\begin{itemize}
	\item two  disjoint  sets of configurations, $S_0$ and $S_1$, %respectively associated with player $0$ and $1$, 
	whose disjoint union $S_0 \cup S_1$ we denote by $S$, and
	\item a relation of %labeled 
		transitions
	 ${\rightarrow} \subseteq S \times %\Lambda \times 
	 				S$. 
\end{itemize}
\end{samepage}

>>>>

{\bf Context} Pushdown systems can be used to model the behavior of recursive programs.
As a result, a variety of 
 program analysis questions can be reduced to decision problems for games on pushdown systems.
They have applications for instance in inter-procedural control-flow analysis of recursive programs \cite{esparza1999automata, reps2005weighted}.

@@

{\bf Parametric pushdown systems}
 Parametric pushdown systems provide a  formalism to
reason about 
recursive programs
making use of parametric constraints. 
Recently, different variants of parametric pushdown systems have been introduced in the literature 
\cite{hague2011parameterised, esparza2016parameterized, fortin2017model} 
however as
parameterized asynchronous shared-memory systems.
These systems consist of a leader pushdown automaton
and arbitrarily many identical contributor pushdown systems, communicating via a shared memory
in the form of a register which can take finitely many values.
This variant have been shown to have applications
to the dataflow analysis of concurrent programs \cite{kahlon2008parameterization}. 
We consider here a different approach to extending pushdown systems with parameters, by allowing them to test equality of the stack content against parameters.
Perhaps a similar model consists in 
pushdown systems 
with transitions that are conditioned by regular conditions on their stack content, which
can in particular
test the
stack content against
specific words. 
They can be used to ensure that some word over the stack alphabet appears in the configurations of a run, but only for a specific value (or, rather, specific regular languages).
They have \textcolor{red}{ phrases trop longues}
been used to establish that CTL$^*$ model checking remains decidable
when the formulas are allowed to include regular predicates on the stack content,
and to obtain 
model checking algorithms for LTL and CTL$^*$ model checking
for
pushdown systems \cite{finkel1997direct}.
Pushdown systems 
with transitions that are conditioned by regular conditions on their stack content
can be viewed as
a special case of stack automata as seen in \cite{hopcroft1969formal}.
Instead
of checking the stack content against regular conditions, or, in a more limited manner,
against specified word values,
we consider
checks against unspecified word values that can be instantiated using parameters.
This allows for example to ensure equality between two stack contents appearing in two distincts configurations in a run.

@@

Another closely related formalism is that of alternating tree automata with nested pebbles. The approach of extending tree automata with pebbles was used in \cite{milo2000typechecking} to show that the XML typechecking problem is decidable, and in \cite{karhumaki2012jewels} to recognize first-order logic on trees. In both cases however, the input trees considered are finite, while 
simulation of plays in parametric pushdown graphs
requires the use of pebble alternating tree automata working on infinite trees
since the stack of a pushdown system is in general unbounded.

@@

@@

By $\Z $ we denote the {\em integers} and by $\N=\{0,1,\ldots\}$ we denote the {\em naturals}.
For every finite alphabet $A$, $A^*$ is the set of finite
words with letters in  $A$, $A^\omega$ is the set of infinite words with letters in $A$, and
 $A^\infty = A^\omega \cup A^*$. We denote the empty word in $A^*$ by $\epsilon$.

 For any two sets $X$ and $S$, let $X^S$ denote the set of all functions from $S$ to $X$.
For any set $S$ let 
$\mathcal{P}(S) = \{ X \mid X \subseteq S \}$
denote the power set of $S$.
%
A {\em partial function} $f$ from a set $X$ to a set $Y$ is a function defined on a subset $C$ of $X$ (possibly $X$ itself) with output values in $Y$. We extend the domain of a partial function to the whole set
by considering it as returning the bottom element $\bot$ when it is undefined.

A {\em graph} $G=(V,E)$ is a finite set of graph vertices $V$ with a set of graph edges $E \subseteq V \times V$. 
For any set $V$, and a given infinite sequence $\pi = v_0 v_1 \ldots \in V^\omega$ of elements of $V$, we define the set 
$\text{\textit{Inf}}(\pi)$ of elements occurring infinitely often in $\pi$ as
$\text{\textit{Inf}}(\pi) = \{ v \in V \mid \forall i, \exists j > i, ~ v_j = v \}$.

@@

@@

\begin{samepage}
An {\em arena} is a tuple $A=(S_0, S_1,{\rightarrow})$  which is composed of
\begin{itemize}
	\item two  disjoint  sets of configurations, $S_0$ and $S_1$, 
	whose disjoint union $S_0 \cup S_1$ we denote by $S$, and
	\item a relation of  
		transitions
	 ${\rightarrow} \subseteq S \times 	S$. 
\end{itemize}
\end{samepage}
